Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-11-01 fstats_pca_lewontin.tex
discussion reordering

\section{Discussion}
Particularly for the analysis of human genetic variation, $F$-statistics  are  a powerful tool to describe population genetic diversity. Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA-plot. This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots. As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using generative models, e.g using a lower number of populations. It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots; differences between the two data summaries are either due to variation on higher PCs, or due to differences in assumptions about normalizations or population groupings. Previous interpretation of PCA in the context of population genetic models have focused on simple models such as trees \citep{cavalli-sforza1975}, homogeneous spatial models \citep{novembre2008a} and discrete-population models \citep{franccois2021}. My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs. One consequence is that the results here are not impacted by sample ascertainment, sample sizes or number of principal components analyzed, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis. From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.

The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized. In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies. In contrast, most data analyses have to grapple with missing data, SNP are often weighted according to their allele frequency and observed, individual-level genotypes are used as the basis of PCA.

The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}. Alternatively, samples with large amounts of missing data are projected onto PCAs computed without missing data \citep{patterson2006}. In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation. These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}.

The normalization of SNPs is similarly a matter of convention. The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable; and the same would hold if SNPs were weighted. The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}.

@@

>>>>

\section{Discussion}
Particularly for the analysis of human genetic variation with a large number of individuals with heterogeneous relationships, $F$-statistics  are  a powerful tool to describe population genetic diversity. Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA-plot. This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots. As PCA is often ran in an early step in data analysis, this may aid in generation of hypotheses that can be more directly evaluated using generative models, typically using a lower number of populations. It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots. In many cases, differences between the two data summaries will be due to variation on higher PCs. In this case, plotting additional PCs, or further subsetting the data to a more local set of populations seems prudent.

The other cause for disagreements between $F$-statistics and PCA are differences in assumptions. The version of PCA I used for my analyses was chosen such that the similarities to $F$-statistics are maximized. In particular, I assume here that i) we have no missing data, ii) SNPs are equally weighted and iii) that individuals can be grouped into populations and iv) we use estimated allele frequencies. In contrast, most data analyses have to grapple with missing data, SNP are often weighted according to their allele frequency and observed, individual-level genotypes are used as the basis of PCA.

The liability of PCA to missing data is a well-studied problem and a number of algorithms for imputing have been proposed \citep[e.g.][]{hastie2015, meisner2017, meisner2021}. Calculating $F$-statistics first is yet another way to impute data for PCA that works well if missingness is low \citep{meisner2021}. In contrast, missing data in $F$-statistics is handled by estimating a standard error using resampling across the genome \citep{patterson2012}, which does not distinguish between biological and sampling variation. These strategies are distinct, but not unique to the relative approaches and a PCA-like decomposition from $F$-statistics is commonly applied using MDS \citep[e.g.][]{fu2016}, and a PCA-based approach to estimating $F$-statistics may provide higher accuracy than naive approaches because it allows using data from related populations.

The normalization of SNPs is similarly a matter of convention. The $F$-statistic framework assumes that each SNP is an identically-distributed (but not independent) random variable; and the same would hold if SNPs were weighted. The drawback for individual $F$-statistics is that this adds a dependency on additional samples (through the mean allele frequency) that may be unwanted for individual $F$-statistics, but could be advantageous for tools that aim to do joint inference from  many $F$-statistics \citep{patterson2012, harney2021}. As genetic differentiation between human populations is low, the  normalization used may matter little in practice \citep{felsenstein1973}, but using consistent normalization between $F$-statistics and PCA-based analysis seems advisable since it removes normalization as a source of errors. .

@@

Previous interpretation of PCA in the context of population genetic models have focused on simple models such as trees \citep{cavalli-sforza1975}, homogeneous spatial models \citep{novembre2008a} and discrete-population models \citep{franccois2021}. My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs. One consequence is that the results here are not impacted by sample ascertainment, sample sizes or number of principal components analyzed, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis. From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data.
